Circumnuclear Media and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-02-10 ms.tex
more edits 5.3

\begin{figure}
  \includegraphics[width=\columnwidth]{profiles.eps}
  \caption{\label{fig:profiles}Radial profiles of the CNM density
    ({\it top}), temperature ({\it middle}), velocity ({\it bottom}),
    and interior X-ray luminosity ({\it bottom}) calculated for a
    sample of galaxies within our grid.  Different colors
    denote different values of the effective wind heating rate,
    $\vwO=1200$ km s$^{-1}$ ({\it blue}), $\vwO$=600 km s$^{-1}$ ({\it
      orange}), and 300 km s$^{-1}$ ({\it green}).  Different
    linestyles correspond to different values of $\Mbh$, $\Mbh=10^6
    \Msun$ ({\it dot-dashed}), $\Mbh=10^7 \Msun$ ({\it solid}), and
    $\Mbh=10^8 \Msun$ ({\it dashed}). Thin lines correspond to cusp
    galaxies ($\Gamma$=0.8) and thick lines correspond to core
    galaxies ($Gamma$=0.1). Squares mark out the location of
    the stagnation radius for each solution.
 }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{rs.eps}
  \caption{\label{fig:stag} \emph{Top panel:} Stagnation radius
    $r_{s}$ in units of the sphere of influence radius $r_{\rm
      soi}$ (eq.~[\ref{eq:rsoi}]) for galaxies in our sample as a
    function of the ratio of the effective wind heating rate $v_{w}$
    to the stellar velocity dispersion $\sigma$.  Green, orange, and
    blue symbols correspond to different values of $v_{w} =$ 300, 600, and 1200 km s$^{-1}$,
    respectively.  Squares correspond to cusp galaxies ($\Gamma = 0.8$), while
    triangles correspond to cores ($\Gamma = 0.1$). The black curves correspond to the
    analytic prediction from equation (\ref{eq:rs2main}), with solid and dashed curves calculated for parameters $\{\Gamma=0.8, n=1\}$ and $\{\Gamma=0.1,n=0.5\}$, respectively.  {\bf BDM: is the x axis here $v_{w}$ or $\vw$ !!?  fold $\eta$ into the y axis. }}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{cooling.eps}
  \caption{\label{fig:cooling} Ratio of the rate of heating to radiative cooling, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$, as a function of radius (solid lines) for each of our solutions shown in Figure \ref{fig:profiles}.  Dashed lines show for comparison the ratio of the cooling to the free-fall timescale $t_{\rm cool}/t_{\rm ff}$.  For high heating rates both ratios are approximately equal at the stagnation radius (marked as squares).  When $\dot{q}_{\rm heat}/\dot{q}_{\rm cool} \lesssim 1$ and/or $t_{\rm cool}/t_{\rm ff} \lesssim 10$, the flow is susceptible thermal instabilities.  {\bf BDM: change y axis labels to qdotheat/qdotcool, tcool/tff}.}
\end{figure}

@@

Appendix \ref{app:windheat} provides a calculation of wind heating rate from stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), as a function of age $\tau_{\star}$ of a stellar population which is formed impulsively (Figs.~\ref{fig:vwimp}, \ref{fig:etaImp}).  At very early times after the starburst ($\tau_{\star} \lesssim 10^{7}$ yr), the wind heating rate exceeds 1000 km s$^{-1}$, while at much later times ($\tau_{\star} \sim t_{\rm h}$) stellar wind heating dominated by main sequence winds is much lower, $v_{\rm w} \sim 50-100 $ km s$^{-1}$.  As we describe in $\S$, the net heating rate produced by stellar winds in the case of quasi-continuous star formation, as representative of the average star formation history of low mass galaxies, can also be significant, $v_{\rm w}^{\star} \gtrsim 1000$ km s$^{-1}$.  Thus stellar winds contribute a potentialy important source of energy as well as mass for the CNM.

@@

Type Ia supernovae (SNe Ia) represent a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  If each Ia supernova injects thermal energy $E_{\rm Ia}$ into the interstellar medium, and the SNe rate (per stellar mass) is given by $R_{\rm   Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm  Ia}$ produces an effective wind heating parameter (equation~[\ref{eq:vw_eff}]) \begin{align} v_{w}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ ergs depends on the efficiency $\epsilon_{\rm Ia}$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.  Hereafter we adopt $\epsilon_{\rm Ia} = 0.4$ as fiducial.

The Ia SN rate $\RateIa$ depends on the age of the stellar population, as it represent the convolution of the star formation rate and the Ia delay time distribution (DTD) divided by the present stellar mass.  In
the limit of impulsive star formation, $\RateIa$ is the DTD evaluated at the time since the star formation episode.  The observationally inferred DTD (Fig. 1 of \citealt{MaozMannucci+:2012a}) has the
approximate functional form \begin{align}
  R_{\rm Ia} =1.7\times 10^{-14}\left(\tau_{\star}/t_{\rm
      h}\right)^{-1.12} M_{\odot}^{-1}\,{\rm yr^{-1}}
\label{eq:DTD}
  \end{align}
  where $\tau_{\star}$ is the time since star formation
  (c.f.~\citealt{Scannapieco&Bildsten05}).

@@

The high value of $v_{w}^{\rm Ia}$ implies that Ia SNe represent an important source of CNM heating.  However, SNe can only be approximated as supplying constant heating throughout space and time if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  We define the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \sim
    38 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim (M_{\rm enc}R_{\rm Ia})^{-1} \sim G/(r\sigma^{2}R_{\rm Ia})$ exceeds the local dynamical timescale $t_{\rm
dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship
(equation~[\ref{eq:Msigma}]).

Assuming that $v_{w}^{\rm Ia} \gg \sigma$ then by substituting $v_{w}^{\rm Ia}$ (eq.~\ref{eq:vIa}) into equation (\ref{eq:rs_simple}) for the stagnation radius, we find that
\begin{eqnarray}
\left.\frac{r_{\rm Ia}}{r_{\rm s}}\right|_{\rm v_{w}^{\rm Ia}} &\approx& 12.4\eta_{0.02} M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{1.68} \nonumber \\
&\approx& 12.4M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{0.38}
\label{eq:rIars}
\end{eqnarray}
The fact that $r_{\rm Ia} \gg r_{\rm s}$ appears to imply that Ia heating can only be approximated as a steady heating source near the stagnation radius for extremely massive SMBHs $M_{\bullet} \gtrsim 10^9 \Msun$ or very young populations ($\tau_{\star} \ll t_{\rm h}$).  However, equation (\ref{eq:rIars}) may underestimate the importance of Ia heating because $r_{\rm Ia}/r_{\rm s}$ will be substantially smaller than its value estimated using $v_{w}^{\rm Ia}$ if the nominal heating rate (without Ia contribution) is much less than this.  Nevertheless, the impact of Ia SN is clearly greater for larger SMBH masses for which $r_{\rm Ia} \lesssim r_{\rm s}$.

Another effect of SN Ia is that they cap the accretion rate onto the SMBH by periodically blowing gas out of the nucleus of low mass galaxies.  Between subsequent supernovae, stars release a gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence Ia SN can in principle dynamically clear out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$, especially for low mass BHs with $\sigma \ll v^{\rm Ia}_{w}$.  Thus, even in cases when $r_{\rm s} > r_{\rm Ia}$, the accretion rate onto low mass SMBHs is limited to the value set by substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]), i.e.  
\begin{eqnarray}
\frac{\dot{M}_{\rm max,Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp}.
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
Note that in general $\dot{M}_{\rm max,Ia}$ exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm max,TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow out' from SNe Ia cannot alone prevent cooling instabilities in low mass galaxies.

{\bf AG: Changed this expression somewhat from that of BDM (slightly
  different scaling relations were used?). I will
  check again! Note that the time dependence of eta~t$^{-1.3}$ is
  folded in above.}

@@

The average spin-down luminosity of millisecond pulsars in the field
is estimated to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$ based
on the ATNF radio pulsar catalog (\citealt{Manchester+05}), in which
case $v_{w} \lesssim 30$ km s$^{-1}$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{w}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the optimistc assumption $\epsilon_{\rm msp} = 1$.  On the other hand, the
binaries giving rise to MSPs may be disocciated by stellar
interactions in the dense nuclear cluster, reducing their numbers as
compared to the field population estimate above.  MSPs are thus unlikely to be an important source of heating in comparison to stellar heating.

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{profiles.eps}
  \caption{\label{fig:profiles}Radial profiles of the CNM density
    ({\it top}), temperature ({\it middle}), and velocity ({\it bottom}),
    calculated for a representative sample of galaxies.  Different colors
    denote different values of the effective wind heating rate,
    $\vwO=1200$ km s$^{-1}$ ({\it blue}), $\vwO$=600 km s$^{-1}$ ({\it
      orange}), and 300 km s$^{-1}$ ({\it green}).  Different
    linestyles correspond to different values of $\Mbh$, $\Mbh=10^6
    \Msun$ ({\it dot-dashed}), $\Mbh=10^7 \Msun$ ({\it solid}), and
    $\Mbh=10^8 \Msun$ ({\it dashed}). Thin lines correspond to cusp
    galaxies ($\Gamma$=0.8) and thick lines correspond to core
    galaxies ($\Gamma$=0.1). Squares mark the location of
    the stagnation radius for each solution {\bf BDM: added colored labels for the $v_w$ value, e.g. a blue 1200 km/s directly to the top figure} .
 }
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{rs.eps}
  \caption{\label{fig:stag} \emph{Top panel:} Stagnation radius
    $r_{s}$ in units of the sphere of influence radius $r_{\rm
      soi}$ (eq.~[\ref{eq:rsoi}]) for galaxies in our sample as a
    function of the ratio of the effective wind heating rate $v_{w}$
    to the stellar velocity dispersion $\sigma$.  Green, orange, and
    blue symbols correspond to different values of $v_{w} =$ 300, 600, and 1200 km s$^{-1}$,
    respectively.  Squares correspond to cusp galaxies ($\Gamma = 0.8$), while
    triangles correspond to cores ($\Gamma = 0.1$). The black curves correspond to the
    analytic prediction from equation (\ref{eq:rs2main}), with solid and dashed curves calculated for parameters $\{\Gamma=0.8, n=1\}$ and $\{\Gamma=0.1,n=0.5\}$, respectively. }
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{cooling.eps}
  \caption{\label{fig:cooling} Ratio of the rate of heating to radiative cooling, $\dot{q}_{\rm heat}/|\dot{q}_{\rm rad}|$, as a function of radius (solid lines) for each of our solutions shown in Figure \ref{fig:profiles}.  Dashed lines show for comparison the ratio of the cooling to the free-fall timescale $t_{\rm cool}/t_{\rm ff}$.  For high heating rates both ratios are approximately equal at the stagnation radius (marked as squares).  When $\dot{q}_{\rm heat}/\dot{q}_{\rm cool} \lesssim 1$ and/or $t_{\rm cool}/t_{\rm ff} \lesssim 10$, the flow is susceptible thermal instabilities.  {\bf BDM: change y axis labels to qdotheat/qdotrad, tcool/tff}.}
\end{figure}

@@

Appendix \ref{app:windheat} provides a calculation of wind heating rate from stellar winds, $v_{\rm w}^{\star}$, and the mass loss parameter, $\eta$ (eq.~[\ref{eq:q}]), as a function of age $\tau_{\star}$ of a stellar population which is formed impulsively (Figs.~\ref{fig:vwimp}, \ref{fig:etaImp}).  At very early times after the starburst ($\tau_{\star} \lesssim 10^{7}$ yr), the wind heating rate exceeds 1000 km s$^{-1}$, while at much later times ($\tau_{\star} \sim t_{\rm h}$) stellar wind heating dominated by main sequence winds is much lower, $v^{\star}_{\rm w} \sim 50-100 $ km s$^{-1}$.  As we describe in $\S$, the net heating rate produced by stellar winds in the case of quasi-continuous star formation, as representative of the average star formation history of low mass galaxies, can also be significant, $v_{\rm w}^{\star} \gtrsim 1000$ km s$^{-1}$.  Thus stellar winds contribute a potentialy important source of energy as well as mass for the CNM.

@@

Type Ia supernovae (SNe Ia) represent a potentially important source of heating, which unlike core collapse SNe is present even in an evolved stellar population.  If each Ia supernova injects thermal energy $E_{\rm Ia}$ into the interstellar medium, and the SNe rate (per stellar mass) is given by $R_{\rm   Ia}$, then the resulting volumetric heating rate $E_{\rm Ia}R_{\rm  Ia}$ produces an effective wind heating parameter (eq.~[\ref{eq:vw_eff}]) \begin{align} v_{w}^{\rm Ia} =\sqrt{\frac{2 t_h R_{\rm Ia}
E_{\rm Ia}}{\eta}} \label{eq:vw_sne}.
\end{align} The thermal energy injected by a Ia SN, $E_{\rm Ia} \simeq
\epsilon_{\rm Ia} 10^{51}$ erg, depends on the efficiency $\epsilon_{\rm Ia}$
with which the initial blast wave energy is converted into bulk or turbulent
motion instead of being lost to radiation.  \cite{Thornton+98}
estimate a radiative efficiency $\epsilon_{\rm Ia} \sim 0.1$,
depending weakly on surrounding density, but \citet{Sharma+14} argues
that $\epsilon_{\rm Ia}$ can be considerably higher, $\sim 0.4$, if
the SNe occur in a hot dilute medium, as may characterize the circumnuclear medium.  Hereafter we adopt $\epsilon_{\rm Ia} = 0.4$ as fiducial.

The Ia SN rate, $\RateIa$, depends on the age of the stellar population, as it represent the convolution of the star formation rate and the Ia delay time distribution (DTD) divided by the present stellar mass.  In
the limit of impulsive star formation, $\RateIa$ is the DTD evaluated at the time since the star formation episode.  The observationally inferred DTD (Fig. 1 of \citealt{MaozMannucci+:2012a}) has the
approximate functional form \begin{align}
  R_{\rm Ia} =1.7\times 10^{-14}\left(\tau_{\star}/t_{\rm
      h}\right)^{-1.12} M_{\odot}^{-1}\,{\rm yr^{-1}}
\label{eq:DTD}
  \end{align}
  where $\tau_{\star}$ is the time since star formation
  (c.f.~\citealt{Scannapieco&Bildsten05}).

@@

The high value of $v_{w}^{\rm Ia}$ implies that Ia SNe represent an important source of CNM heating.  However, SNe can only be approximated as supplying heating which is spatially and temporally homogeneous if the rate of SNe is rapid compared to the inflow time at the radius of interest (\citealt{ShcherbakovWong+:2014a}).  We define the ``Ia radius"
  \begin{align}
    r_{\rm Ia} \sim \left(\frac{G}{R_{\rm Ia}\sigma}\right)^{1/2} \sim
    38 M_{\bullet,8}^{-0.1}(\tau_{\star}/t_{\rm h})^{0.56}\,{\rm pc}
    \label{eq:rIa}
  \end{align}
as the location exterior of which the time interval between subsequent supernovae $\tau_{\rm Ia} \sim (M_{\rm enc}R_{\rm Ia})^{-1} \sim G/(r\sigma^{2}R_{\rm Ia})$ exceeds the local dynamical timescale $t_{\rm
dyn} \sim r/\sigma$, where we again adopt the Ia rate for an old stellar population and in the final equality we estimate the velocity dispersion using the $M_{\bullet}-\sigma$ relationship
(eq.~[\ref{eq:Msigma}]).

Assuming that $v_{w}^{\rm Ia} \gg \sigma$ then by substituting $v_{w}^{\rm Ia}$ (eq.~\ref{eq:vIa}) into equation (\ref{eq:rs_simple}) for the stagnation radius, we find that
\begin{eqnarray}
\left.\frac{r_{\rm Ia}}{r_{\rm s}}\right|_{\rm v_{w}^{\rm Ia}} &\approx& 12.4\eta_{0.02} M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{1.68} \nonumber \\
&\approx& 12.4M_{\bullet,8}^{-1.1}(\epsilon_{\rm Ia}/0.4)^{-1}(\tau_{\star}/t_{\rm h})^{0.38}
\label{eq:rIars}
\end{eqnarray}
If $r_{\rm Ia} \gg r_{\rm s}$, then Ia heating can only be approximated as a steady heating source near the stagnation radius for extremely massive SMBHs with $M_{\bullet} \gtrsim 10^9 \Msun$ or in the case of a very young stellar population ($\tau_{\star} \ll t_{\rm h}$).  On the other hand, equation (\ref{eq:rIars}) may underestimate the importance of Ia heating because $r_{\rm Ia}/r_{\rm s}$ will be substantially smaller than its value estimated using $v_{w}^{\rm Ia}$ if the heating rate without SN Ia is less than this.

Even if SN Ia are rare near the stagnation radius, they may cap the SMBH accretion rate by periodically blowing gas out of the nucleus of low mass galaxies.  Between subsequent supernovae, stars release a gaseous mass $M_{\rm g} \approx \eta M_{\star}\tau_{\rm Ia}/t_{\rm h}$ interior to the Ia radius, which is gravitationally bound to the SMBH by an energy $E_{\rm bind} \sim M_{\rm g}\sigma^{2}$.  Thus, according to the above definitions, $E_{\rm Ia}/E_{\rm bind} \sim (v_{w}^{\rm Ia})^{2}/2\sigma^{2} \gtrsim 1$.  Hence, SN Ia are in principle capable of dynamically clearing out gas from radii $\sim r_{\rm Ia} \gtrsim r_{\rm s}$, especially for low mass BHs with $\sigma \ll v^{\rm Ia}_{w}$.  Thus, even in cases when $r_{\rm s} > r_{\rm Ia}$, the accretion rate is limited to the value, 
\begin{eqnarray}
\frac{\dot{M}_{\rm max,Ia}}{\dot{M}_{\rm edd}} &\approx& \frac{\eta M_{\bullet}}{\dot{M}_{\rm edd} t_{\rm h}}\left(\frac{R_{\rm Ia}}{R_{\rm soi}}\right)^{2-\Gamma} \approx \nonumber \\
 && \begin{cases}
    4.5 \times 10^{-4} M_{\bullet,8}^{-1.33}(\tau_{\star}/t_{\rm h})^{-0.2}
   & \text{core} \\
    2.2 \times 10^{-4} \Mbheight^{-0.84}(\tau_{\star}/t_{\rm h})^{-0.6}   & \text{cusp},
  \end{cases}
  \label{eq:eddr_Ia}
\end{eqnarray}
obtained substituting the Ia radius $r_{\rm Ia}$ (eq.~[\ref{eq:rIa}]) for $r_{\rm s}$ in the derivation leading to our estimate of $\dot{M}$ (eq.~[$\ref{eq:eddr_analytic}$]).  Note that $\dot{M}_{\rm max,Ia}$ usually exceeds the maximum accretion rate for a thermally stable flow $\dot{M}_{\rm max,TI}$ (eq.~[\ref{eq:Mdotmax}]), implying that dynamical `blow-out' from SNe Ia cannot alone prevent cooling instabilities in low mass galaxies.  Also note that this maximum Ia accretion rate only applies if the gas within the CNM is supplied by local stellar mass loss, since the quantity of gaseous mass can be much greater if is fed externally, e.g. due to a galactic merger.

@@

The average spin-down luminosity of millisecond pulsars in the field
is estimated to be $\bar{L}_{\rm sd} \sim 10^{34}$ erg s$^{-1}$ based
on the ATNF radio pulsar catalog (\citealt{Manchester+05}), in which
case $v_{w}^{\rm MSP} \lesssim 30$ km s$^{-1}$.  For higher spin-down
luminosities, $L_{\rm sd}\simeq 10^{35}$ ergs s$^{-1}$ characteristic
of some Fermi-detected pulsars, then the higher value of $v_{w}^{\rm MSP}
\lesssim 300$ km s$^{-1}$ makes MSP heating in principle important
under the optimistc assumption $\epsilon_{\rm msp} = 1$.  On the other hand, the
binaries giving rise to MSPs may be disocciated by stellar
interactions in the dense nuclear cluster, reducing their numbers as
compared to the field population estimate above.  MSPs are thus unlikely to be an important source of heating in comparison to stellar heating.
